MAESTROeX: A Massively Parallel Low Mach Number Astrophysical Solver

### 2019-03-09 MAESTROeX_I/paper.tex
updates/simplifications.  need to talk about projection splitting still

\correspondingauthor{MAESTROeX}
\email{DFan@lbl.gov}

@@

Previously we adopted an approach where we split the velocity into a base state component, $w_0$, and a local velocity $\Ubt$, so that
\begin{equation}
\Ub = w_0(r,t)\eb_r + \Ubt(\xb,t).
\end{equation}
We used $w_0$ to provide an estimate of the base state density evolution over a time step.
In our new simplified approach, we instead use a predictor-corrector approach for evolving the base state.

@@

\subsection{A Possible Solution}
We could choose to no longer advect the base states for both predictor and corrector steps. This would solve the problems with computing state values at cell edges. Instead, we propose to only update the base states $\rho_0$ and $p_0$ by computing the average of the full state values after advecting the full states. This would result in the following changes in the algorithm from Paper V:
%
\begin{enumerate}
\item Steps 4 and 8 will be greatly simplified, leading to a more efficient algorithm.
\item There is no longer a need for $w_0$, so we can use the full velocity $\Ub$ instead of $\Ubt$. As a consequence, $\eta$ and $\psi$ also do not need to be computed. This has the added advantage of saving some memory storage by removing many of the terms needed when we decided to split terms into its base state and perturbation state in our previous implementation.
\item By placing the base states exactly at cell-centers, we eliminate interpolation errors and simplify the existing code.
\end{enumerate}
%

@@

At the beginning of each time step we have the state,
$(\Ub,\rho X_k,\rho h,\rho_0,p_0)^n$, and $\pi^{n-\myhalf}$.
At any time, the associated density, composition, and enthalpy can be trivially computed using, e.g.,
\begin{equation}
\rho^n = \sum_k(\rho X_k)^n, \quad
X_k^n = (\rho X_k)^n / \rho^n, \quad
h^n = (\rho h)^n / \rho^n.
\end{equation}
Temperature is computed using the equation of state, e.g.,
\footnote{As described in Paper V, for planar problems we compute temperature using $h$ instead of $p_0$, since we have successfully developed planar volume discrepancy schemes to effectively couple the enthalpy to the rest of the solution.  We are still exploring this option for spherical stars.}
\begin{equation}
T^n = T(\rho,p_0,X_k)^n.
\end{equation}
Similarly, $\gammaonebar$ and $\beta_0$ are computed using $\rho,p_0$, and $X_k$ (see Appendix A of Paper I and Appendix C of Paper III for details).

@@

Using equation (\ref{eq:momentum}), 
we compute time-centered edge velocities, $\uadvonedag$, using
$\Ub^n$.  The $\dagger$ superscript refers to the
fact that the predicted velocity field does not satisfy the divergence
constraint.  We then construct $\uadvone$ from $\uadvonedag$
using a MAC projection.
We note that $\uadvone$ satisfies the divergence constraint
\begin{equation}
\nabla \cdot \left(\beta_0^n \uadvone\right) = \beta_0^n \left(S^{\nph,\star} - \frac{1}{\gammaonebar^np_0^n}\psi^{n-\myhalf} \right).
\end{equation}

It should also be noted that we have found it better to approximate 
the temporal changes in base pressure as 
\begin{equation}
\frac{\partial p_0}{\partial t} \approx \psi = \eta_{\rho} g
\end{equation}
instead of using the lagged pressure changes from previous time $t^{n-1}$ 
in the predictor step.

@@

\item Evolve $(\rho h)^{(1)} \rightarrow (\rho h)^{(2),\star}$.
\begin{description}
\item For planar geometry,
\begin{eqnarray}
(\rho h)^{(2),\star}
&=& (\rho h)^{(1)} \nonumber \\
&&- \dt^n \left\{ \nabla \cdot \left[ \uadvone (\rho h)^{\nph,\pred} \right] \right\} \nonumber \\
&& + \dt^n \left(\uadvone \cdot \eb_r\right) \left(\frac{\partial p_0}{\partial r} \right)^{n} \nonumber \\
&& + \dt^n \psi^{n-\myhalf},
\end{eqnarray}

\item For spherical geometry,
\begin{eqnarray}
(\rho h)^{(2),\star}
&=& (\rho h)^{(1)} \nonumber \\
&&- \dt^n \left\{ \nabla \cdot \left[ \uadvone (\rho h)^{\nph,\pred} \right] \right\} \nonumber \\
&& + \dt^n \left \{ \nabla \cdot \left (\uadvone p_0^{n} \right ) - p_0^{n} \nabla \cdot \uadvone \right \} \nonumber \\
&& + \dt^n \psi^{n-\myhalf},
\end{eqnarray}
\end{description}

@@

The procedure to construct $\uadvtwodag$ is identical to the procedure
for computing $\uadvonedag$ in {\bf Step 3}, but uses
the updated value $S^{\nph}$ rather than $S^{\nph,\star}$.
Also, the temporal changes of $p_0$ are now computed directly instead of 
approximated using $\psi$.
We note that $\uadvtwo$ satisfies the divergence constraint
\begin{equation}
\nabla \cdot \left(\beta_0^{\nph} \uadvtwo\right) =
\beta_0^{\nph} \left(S^{\nph} - \frac{1}{\gammaonebar^{n+\myhalf}p_0^{n+1,\star}}\frac{p_0^{n+1,\star}-p_0^n}{\Delta t^n}\right),
\end{equation}
where
\begin{equation}
\beta_0^{\nph} = \frac{ \beta_0^n +  \beta_0^{n+1,\star} }{2},
\qquad
\gammaonebar^{\nph} = \frac{ \gammaonebar^n +  \gammaonebar^{n+1,\star} }{2}.
\end{equation}

@@

\item Evolve $(\rho X_k)^{(1)} \rightarrow (\rho X_k)^{(2)}$ using
\begin{equation}
(\rho X_k)^{(2)} = (\rho X_k)^{(1)}
- \dt \left\{ \nabla \cdot \left[\uadvtwo (\rho X_k)^{\nph,\pred} \right] \right\},
\end{equation}
\begin{equation}
\rho^{(2)} = \sum_k (\rho X_k)^{(2)},
\qquad
X_k^{(2)} = (\rho X_k)^{(2)} / \rho^{(2)}.
\end{equation}

@@

\item Update the enthalpy.  This step is identical to {\bf Step 4D} except we use
  the updated values $\uadvtwo, \rho_0^{n+1}, (\rho h)_0^{n+1}$, and $p_0^{n+\myhalf}$
  rather than
  $\uadvone, \rho_0^{n+1,\star}, (\rho h)_0^{n+1,\star}$, and $p_0^n$.
  Also, the temporal changes of $p_0$ are now computed directly instead of 
  approximated using $\psi$.
  In particular:
\MarginPar{Again, we may not need $\rho_0$ and $(\rho h)_0$.}

@@

\item Evolve $(\rho h)^{(1)} \rightarrow (\rho h)^{(2)}$.
\begin{description}
\item For planar geometry,
\begin{eqnarray}
(\rho h)^{(2)}
&=& (\rho h)^{(1)} - \dt \left\{ \nabla \cdot \left[ \uadvtwo (\rho h)^{\nph,\pred} \right] \right\} \nonumber \\
&& + \dt \left(\uadvtwo \cdot \eb_r\right) \left(\frac{\partial p_0}{\partial r} \right)^\nph \nonumber \\
&& + \dt \left( \frac{p_0^{n+1} - p_0^n}{\dt^n} \right),
\end{eqnarray}

\item For spherical geometry,
\begin{eqnarray}
(\rho h)^{(2)}
&=& (\rho h)^{(1)} - \dt \left\{ \nabla \cdot \left[ \uadvtwo (\rho h)^{\nph,\pred} \right] \right\} \nonumber \\
&& + \dt \left[ \nabla \cdot \left (\uadvtwo p_0^{\nph} \right ) - p_0^{\nph} \nabla \cdot \uadvtwo \right] \nonumber \\
&& + \dt \left( \frac{p_0^{n+1} - p_0^n}{\dt^n} \right),
\end{eqnarray}
\end{description}
where $p_0^\nph$ is defined as $p_0^\nph = (p_0^n+p_0^{n+1})/2$.

@@

\item Compute $\psi^{\nph}$: 
\begin{equation}
\psi_j^{\nph} = \eta_{\rho,j}^{\nph} g.
\end{equation}

@@

First, we compute the time-centered edge velocities, $\Ub^{\nph,\pred}$.
Then, we define
\begin{equation}
\rho^\nph = \frac{\rho^n + \rho^{n+1}}{2}, \qquad \rho_0^\nph = \frac{\rho_0^n + \rho_0^{n+1}}{2}.
\end{equation}
We update the velocity field $\Ub^n$ to $\Ub^{n+1,\dagger}$ by discretizing
equation (\ref{eq:momentum}) as
\begin{eqnarray}
\Ub^{n+1,\dagger}
&=& \Ub^n - \dt \left[\uadvtwo \cdot \nabla \Ub^{\nph,\pred} \right] \nonumber \\
&& - \dt \left[ \frac{1}{\rho^\nph} \mathbf{G} \pi^\nmh + \frac{\left(\rho^\nph-\rho_0^\nph\right)}{\rho^\nph} g^{\nph} \eb_r \right],\nonumber \\
\end{eqnarray}
where $\mathbf{G}$ approximates a cell-centered gradient from nodal
data.  Again, the $\dagger$ superscript refers
to the fact that the updated velocity does not satisfy the divergence
constraint.

Finally, we use an approximate nodal projection to define $\Ub^{n+1}$
from $\Ub^{n+1,\dagger},$  such that $\Ub^{n+1}$ approximately
satisfies
\begin{equation}
\nabla \cdot \left(\beta_0^{n+1} \Ub^{n+1} \right)
= \beta_0^{n+1} S^{n+1}.
\end{equation}
As part of the projection we also define the new-time perturbational pressure,
$\pi^\nph.$  This projection necessarily differs from the MAC projection used in
{\bf Step 3} and {\bf Step 7} because the velocities in those steps are defined
on edges and $\Ub^{n+1}$ is defined at cell centers, requiring different divergence
and gradient operators.  Details of the approximate projection are given in Paper III.

@@

%--------------------------------------------------------------------------
% STEP 0
%--------------------------------------------------------------------------
{\em Initialization}
\sout{
The initialization step only occurs at the beginning of the simulation.
The initialization step computes initial values for $\pi^{-\myhalf}$, , $\eta_{\rho}^{-\myhalf}$, 
$\psi^{-\myhalf}$, $S^0$, and $S^1$ for use in time stepping routine.  
We pass through {\bf Step 1} -- {\bf Step 11}, but only keep these values.
The initial values for $\Ub^0, \rho^0, (\rho h)^0, X_k^0, T^0,
\rho_0^0, p_0^0$, and $\gammaonebar^0$ are specified from the problem-dependent
initial conditions.  The initial time step, $\dt^0$, is computed as in
Paper III.
}

>>>>

\correspondingauthor{Doreen Fan; DFan@lbl.gov}

@@

Previously we adopted an approach where we split the velocity into a base state component, $w_0(r,t)$, 
and a local velocity $\Ubt(\xb,t)$, so that
\begin{equation}
\Ub = \Ubt(\xb,t) + w_0(r,t)\eb_r,
\end{equation}
We used $w_0$ to provide an estimate of the base state density evolution over a time step.
In our new simplified approach, we instead use a predictor-corrector approach for evolving the base state.

@@

\subsection{A Possible Solution}
We could choose to no longer advect the base states for both predictor and corrector steps. This would solve the problems with computing state values at cell edges. Instead, we propose to only update the base states $\rho_0$ and $p_0$ by computing the average of the full state values after advecting the full states. This would result in the following changes in the algorithm from Paper V:
%
\begin{enumerate}
\item Steps 4 and 8 will be greatly simplified, leading to a more efficient algorithm.
\item \sout{There is no longer a need for $w_0$, so we can use the full velocity $\Ub$ instead of $\Ubt$. As a consequence, $\eta$ does not need to be computed. This has the added advantage of saving some memory storage by removing many of the terms needed when we decided to split terms into its base state and perturbation state in our previous implementation.} We split the velocity for the projection to handle the cutoff.
\item By placing the base states exactly at cell-centers, we eliminate interpolation errors and simplify the existing code.
\end{enumerate}
%

@@

At the beginning of each time step we have the cell-centered state,
$(\Ub,\rho X_k,\rho h,\rho_0,p_0)^n$, and nodal state, $\pi^{n-\myhalf}$.
At any time, the associated density, composition, and enthalpy can be trivially computed using, e.g.,
\begin{equation}
\rho^n = \sum_k(\rho X_k)^n, \quad
X_k^n = (\rho X_k)^n / \rho^n, \quad
h^n = (\rho h)^n / \rho^n.
\end{equation}
Temperature is computed using the equation of state, e.g.,
\footnote{As described in Paper V, for planar problems we compute temperature using $h$ instead of $p_0$, since we have successfully developed planar volume discrepancy schemes to effectively couple the enthalpy to the rest of the solution.  We are still exploring this option for spherical stars.}
\begin{equation}
T = T(\rho,p_0,X_k),
\end{equation}
and ($\gammaonebar,\beta_0)$ are computed from $(\rho,p_0,X_k)$ (see Appendix A of Paper I and Appendix C of Paper III for details).

@@

Using equation (\ref{eq:momentum}), 
we compute time-centered edge velocities, $\uadvonedag$, using
$\Ub^n$.  The $\dagger$ superscript refers to the
fact that the predicted velocity field does not satisfy the divergence
constraint.  We then construct $\uadvone$ from $\uadvonedag$
using a MAC projection.
We note that $\uadvone$ satisfies the divergence constraint
\begin{equation}
\nabla \cdot \left(\beta_0^n \uadvone\right) = \beta_0^n \left(S^{\nph,\star} - \frac{1}{\gammaonebar^np_0^n}\frac{\partial p_0}{\partial t} \right).
\end{equation}
In practice we solve for.

@@

\item Evolve $(\rho h)^{(1)} \rightarrow (\rho h)^{(2),\star}$.
\begin{description}
\item For planar geometry,
\begin{eqnarray}
(\rho h)^{(2),\star}
&=& (\rho h)^{(1)} \nonumber \\
&&- \dt^n \left\{ \nabla \cdot \left[ \uadvone (\rho h)^{\nph,\pred} \right] \right\} \nonumber \\
&& + \dt^n \left(\frac{p_0^n - p_0^{n-1}}{\dt^{n-1}}\right) \nonumber \\
&& + \dt^n \left(\uadvone \cdot \eb_r\right) \left(\frac{\partial p_0}{\partial r} \right)^{n}
\end{eqnarray}

\item For spherical geometry,
\begin{eqnarray}
(\rho h)^{(2),\star}
&=& (\rho h)^{(1)} \nonumber \\
&&- \dt^n \left\{ \nabla \cdot \left[ \uadvone (\rho h)^{\nph,\pred} \right] \right\} \nonumber \\
&& + \dt^n \left(\frac{p_0^n - p_0^{n-1}}{\dt^{n-1}}\right) \nonumber \\
&& + \dt^n \left \{ \nabla \cdot \left (\uadvone p_0^{n} \right ) - p_0^{n} \nabla \cdot \uadvone \right \}
\end{eqnarray}
\end{description}

@@

The procedure to construct $\uadvtwodag$ is identical to the procedure
for computing $\uadvonedag$ in {\bf Step 3}, but uses
the updated value $S^{\nph}$ rather than $S^{\nph,\star}$.
We note that $\uadvtwo$ satisfies the divergence constraint
\begin{equation}
\nabla \cdot \left(\beta_0^{\nph} \uadvtwo\right) =
\beta_0^{\nph} \left(S^{\nph} - \frac{1}{\gammaonebar^{n+\myhalf}p_0^{n+\myhalf}}\frac{\partial p_0}{\partial t}\right),
\end{equation}
where
\begin{equation}
\beta_0^{\nph} = \frac{ \beta_0^n +  \beta_0^{n+1,\star} }{2},
\qquad
\gammaonebar^{\nph} = \frac{ \gammaonebar^n +  \gammaonebar^{n+1,\star} }{2}.
\qquad
p_0^{\nph} = \frac{ p_0^n + p_0^{n+1,\star} }{2},
\end{equation}

@@

\item Evolve $(\rho X_k)^{(1)} \rightarrow (\rho X_k)^{(2)}$ using
\begin{equation}
(\rho X_k)^{(2)} = (\rho X_k)^{(1)}
- \dt \left\{ \nabla \cdot \left[\uadvtwo (\rho X_k)^{\nph,\pred} \right] \right\},
\end{equation}

@@

\item Update the enthalpy.  This step is identical to {\bf Step 4D} except we use
  the updated values $\uadvtwo, \rho_0^{n+1}, (\rho h)_0^{n+1}$, and $p_0^{n+\myhalf}$
  rather than
  $\uadvone, \rho_0^{n+1,\star}, (\rho h)_0^{n+1,\star}$, and $p_0^n$.
  In particular:
\MarginPar{Again, we may not need $\rho_0$ and $(\rho h)_0$.}

@@

\item Evolve $(\rho h)^{(1)} \rightarrow (\rho h)^{(2)}$.
\begin{description}
\item For planar geometry,
\begin{eqnarray}
(\rho h)^{(2)}
&=& (\rho h)^{(1)} - \dt \left\{ \nabla \cdot \left[ \uadvtwo (\rho h)^{\nph,\pred} \right] \right\} \nonumber \\
&& + \dt \left( \frac{p_0^{n+1} - p_0^n}{\dt} \right)\nonumber \\
&& + \dt \left(\uadvtwo \cdot \eb_r\right) \left(\frac{\partial p_0}{\partial r} \right)^\nph,
\end{eqnarray}

\item For spherical geometry,
\begin{eqnarray}
(\rho h)^{(2)}
&=& (\rho h)^{(1)} - \dt \left\{ \nabla \cdot \left[ \uadvtwo (\rho h)^{\nph,\pred} \right] \right\} \nonumber \\
&& + \dt \left( \frac{p_0^{n+1} - p_0^n}{\dt} \right)\nonumber \\
&& + \dt \left[ \nabla \cdot \left (\uadvtwo p_0^{\nph} \right ) - p_0^{\nph} \nabla \cdot \uadvtwo \right],
\end{eqnarray}
\end{description}
where $p_0^\nph$ is defined as $p_0^\nph = (p_0^n+p_0^{n+1})/2$.

@@

@@

First, we compute the time-centered edge velocities, $\Ub^{\nph,\pred}$.
Then, we define
\begin{equation}
\rho^\nph = \frac{\rho^n + \rho^{n+1}}{2}, \qquad \rho_0^\nph = \frac{\rho_0^n + \rho_0^{n+1}}{2}.
\end{equation}
We update the velocity field $\Ub^n$ to $\Ub^{n+1,\dagger}$ by discretizing
equation (\ref{eq:momentum}) as
\begin{equation}
\Ub^{n+1,\dagger}
= \Ub^n - \dt \left[\uadvtwo \cdot \nabla \Ub^{\nph,\pred} \right]
 - \dt \left[ \frac{1}{\rho^\nph} \nabla \pi^\nmh + \frac{\left(\rho^\nph-\rho_0^\nph\right)}{\rho^\nph} g^{\nph} \eb_r \right],
\end{equation}
Again, the $\dagger$ superscript refers
to the fact that the updated velocity does not satisfy the divergence
constraint.

Finally, we use an approximate nodal projection to define $\Ub^{n+1}$
from $\Ub^{n+1,\dagger},$  such that $\Ub^{n+1}$ approximately
satisfies
\begin{equation}
\nabla \cdot \left(\beta_0^{n+1} \Ub^{n+1} \right) = \beta_0^{n+1} \left( S^{n+1} - \frac{1}{\gammaonebar^{n+1}p_0^{n+1}}\frac{\partial p_0}{\partial t} \right).
\end{equation}
As part of the projection we also define the new-time perturbational pressure,
$\pi^\nph.$  This projection necessarily differs from the MAC projection used in
{\bf Step 3} and {\bf Step 7} because the velocities in those steps are defined
on edges and $\Ub^{n+1}$ is defined at cell centers, requiring different divergence
and gradient operators.  Details of the approximate projection are given in Paper III.

@@

%--------------------------------------------------------------------------
% STEP 0
%--------------------------------------------------------------------------
{\em Initialization}
\sout{
The initialization step only occurs at the beginning of the simulation.
The initialization step computes initial values for 
$\pi^{-\myhalf}, \eta_{\rho}^{-\myhalf}, S^0$, and $S^1$ for use in time stepping routine.  
We pass through {\bf Step 1} -- {\bf Step 11}, but only keep these values.
The initial values for $\Ub^0, \rho^0, (\rho h)^0, X_k^0, T^0,
\rho_0^0, p_0^0$, and $\gammaonebar^0$ are specified from the problem-dependent
initial conditions.  The initial time step, $\dt^0$, is computed as in
Paper III.
}
